---
title: "How Are You Going to Die? A Final Report"
author: "Andrew Zilnicki"
date: "2023-12-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
I got this from data.gov and the population data from the 2010 census.
<br> 
[Link to
Dataset](https://catalog.data.gov/dataset/nchs-leading-causes-of-death-united-states/resource/9096aa3c-0d4b-42f1-bb01-284816d92a15)<br> 

The leading causes of death in the United States are important indicators of the nation's health care system. This information is crucial for policymakers, healthcare professionals, and the general public to develop targeted improvements, allocate resources effectively, and ultimately improve public health outcomes.

My analysis is based on a comprehensive dataset sourced from the National Center for Health Statistics, part of the CDC, and the U.S. Census. Manipulating the data to use statistics allows us to summarize and visualize the key characteristics of the data set, providing an initial understanding of the data. This report can help people by giving them knowledge about health-related issues in their state, and which states offer better healthcare options.

# Questions

**Domain Question**

What are the top causes of deaths by number and by state?

**Other Questions**

1.  Does using the Age-adjusted death rate show different information
    about deaths per state?
2.  Does using the Age-adjusted death rate show different information
    about deaths per cause?
3.  Does using the per-capita death rate show different information
    about deaths per state?
4.  In which states is cancer most deadly using the per-capita death rate?
5.  Which states have the highest rates of suicide using the per-capita death rate?

# Here are the packages I used

```{r message=FALSE}
library(DT) # used for data tables
library(usmap) # used for US map generation
library(dplyr) # used for data manipulation
library(ggplot2) # used for data visualization
```

# Unfiltered Data

This data set is from Data.gov. First, I had to download the data from
Data.gov. The data was downloaded via a CSV file. Once I created an R
Markdown file I uploaded the CSV file into a new data frame called
total_data.

```{r}
total_data <- read.csv("/Users/andrew/DATA 2401/Final project/NCHS_-_Leading_Causes_of_Death__United_States.csv")
glimpse(total_data)
```

## Data Key Terms:

**Year:** Year of death. <br> **X113 Cause Name:** Detailed name for
cause of death. <br> **Cause Name:** Short name for cause of death. <br>
**State:** State where death occurred. <br> **Deaths:** Total deaths for
that death and cause in a particular year. <br> **Age adjusted Death
Rate:** Formula determined by total expected number of deaths divided by
the sum of the standard population and the resulting quotient is
multiplied by 1,000,000. This is the death rate that would have existed
if the population under study had the same age distribution as the
"standard" population. This is useful for summary measures adjusted for
differences in age distributions.

# Data Analysis

To start analyzing data we group our data by state name and cause of
death. We can then plot the data of total deaths by each cause. To make
it easier to view we organize the causes from descending order.

```{r, echo=FALSE, message=FALSE}
result <- total_data %>%
  group_by(Cause.Name, State) %>%
  summarize(TotalDeaths = sum(Deaths)) 
```

```{r, echo=FALSE}
result %>%
  filter(State == "United States" & Cause.Name != "All causes") %>%
  arrange(desc(TotalDeaths)) %>%
  ggplot(aes(x = reorder(Cause.Name, -TotalDeaths), y = TotalDeaths, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Deaths by Cause in the United States",
       x = "Cause Name",
       y = "Total Deaths",
       fill = "Cause Name") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Now, we can filter our data by total deaths for each state ordering from the
least deaths to the most deaths for easy viewing.

```{r, echo=FALSE}
result %>%
  filter(Cause.Name == "All causes" & State != "United States") %>%
  arrange(desc(TotalDeaths)) %>%
  ggplot(aes(x = reorder(State, -TotalDeaths), y = TotalDeaths, fill = State)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flip the plot for a horizontal bar chart
  labs(title = "Total Deaths by State for All Causes",
       x = "Total Deaths",
       y = "State",
       fill = "State") +
  theme_minimal()
```

That's a lot of information. To make it easier to read let's filter it down to the top 5 causes and the top 10 states.

```{r message=FALSE, echo=FALSE}
filtered_data <- result %>%
  filter(
    Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
    State %in% c("California", "Florida", "Texas", "New York", "Pennsylvania", "Ohio", "Illinois", "Michigan", "North Carolina", "New Jersey")
  )
```

```{r, echo=FALSE}
ggplot(filtered_data, aes(x = reorder(State, -TotalDeaths), y = TotalDeaths, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Deaths by Cause in Selected States",
       x = "Total Deaths",
       y = "State",
       fill = "Cause Name") +
  theme_minimal()
```

This shows that California has far more deaths than even the second
closest state of Florida. California has triple the number of deaths
from the top 5 causes once you look at North Carolina or New Jersey.

# A better way to look at causes?

It might look like some states have far more deaths from illness, but
these states also have the largest populations. <br>
What if we used the Age-adjusted data? <br>
Would we get different results for states and causes?

```{r, message=FALSE, echo=FALSE}
new_filtered_data <- total_data %>%
  filter(Cause.Name != "All causes" & State != "United States") %>%
  group_by(Cause.Name, State) %>%
  summarise(TotalAgeAdjustedDeathRate = sum(Age.adjusted.Death.Rate))
```

```{r, message=FALSE, echo=FALSE}
ggplot(new_filtered_data, aes(x = reorder(Cause.Name, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Age-Adjusted Death Rate by Cause Name",
       x = "Total Age-Adjusted Death Rate",
       y = "Cause Name",
       fill = "Cause Name") +
  theme_minimal()
```

This shows that the causes of death remain in the same order of total numbers of deaths.
Let us see if the states change in the graph below.

```{r message=FALSE, echo=FALSE}
ggplot(new_filtered_data, aes(x = reorder(State, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = State)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Age-Adjusted Death Rate by State",
       x = "Total Age-Adjusted Death Rate",
       y = "State",
       fill = "State") +
  theme_minimal()
```

You can see that the order of states have changed. Now, let's look at a table showing the top 5 Age-Adjusted Death Causes and the Top 10 states

```{r message=FALSE, echo=FALSE}
top_causes_states <- new_filtered_data %>%
  filter(Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
           State %in% c("Mississippi", "Kentucky", "Oklahoma", "West Virginia", "Alabama",
                         "Louisiana", "Arkansas", "Tennessee", "Missouri", "South Carolina"))

datatable(top_causes_states, options = list(pageLength = 10))
```

Here is a graph to make visualizing this data easier.

```{r message=FALSE, echo=FALSE}
ggplot(top_causes_states, aes(x = reorder(State, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 5 Age-Adjusted Death Causes by Top 10 States",
       x = "Total Age-Adjusted Death Rate",
       y = "State",
       fill = "Cause Name") +
  theme_minimal()
```

# Using per capita

Another way to determine overall health and death rate is to use each state's population. Let's look at the deaths using population per capita to see if that changes the results. <br>
I am going to pick 2010 and load population data from the census.

```{r message=FALSE}
# Creating a data frame with state populations from the 2010 Census
state_populations_2010 <- data.frame(
  State = c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado",
    "Connecticut", "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho",
    "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana",
    "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",
    "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey",
    "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma",
    "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
    "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia",
    "Wisconsin", "Wyoming"
  ),
  Population_2010 = c(
    4779736, 710231, 6392017, 2915918, 37253956, 5029196,
    3574097, 897934, 601723, 18801310, 9687653, 1360301, 1567582,
    12830632, 6483802, 3046355, 2853118, 4339367, 4533372,
    1328361, 5773552, 6547629, 9883640, 5303925, 2967297,
    5988927, 989415, 1826341, 2700551, 1316470, 8791894,
    2059179, 19378102, 9535483, 672591, 11536504, 3751351,
    3831074, 12702379, 1052567, 4625364, 814180, 6346105,
    25145561, 2763885, 625741, 8001024, 6724540, 1852994,
    5686986, 563626
  )
)
```

```{r message=FALSE, echo=FALSE}
total_data_2010 <- total_data %>% 
  filter(Year == 2010) %>% 
  select(Year, Cause.Name, State, Deaths, Age.adjusted.Death.Rate)

merged_data <- left_join(total_data_2010, state_populations_2010, by = "State")

merged_data <- merged_data %>%
  mutate(Deaths_Per_Capita = Deaths / Population_2010 * 1000)  # Multiplying by 1000 to get deaths per 1000 people
```

This map shows the deaths per 1,000 people for each state. The lower table gives the numbers for the top 10 states.

```{r message=FALSE, echo=FALSE}
merged_data$state <- tolower(merged_data$State)
plot_usmap(data = merged_data, values = "Deaths_Per_Capita", color = "lightblue") +
  labs(title = "Deaths Per Capita by State in 2010",
       subtitle = "Color-coded map based on Deaths Per Capita (per 1,000 people)") +
  theme(legend.position = "bottom", legend.box = "horizontal")

top_merged_data <- merged_data %>%
  filter(Cause.Name == "All causes") %>%
  arrange(desc(Deaths_Per_Capita)) %>%
  top_n(10, Deaths_Per_Capita)

print(top_merged_data[, c("State", "Deaths_Per_Capita")])
```

This map shows the deaths per 1,000 people for people who died because of cancer for each state. The lower table gives the numbers for the top 10 states.

```{r message=FALSE, echo=FALSE}
cancer_data <- merged_data %>%
  filter(Cause.Name == "Cancer")

cancer_data$state <- tolower(cancer_data$State)
  plot_usmap(data = cancer_data, values = "Deaths_Per_Capita", color = "lightblue") +
  labs(title = "Cancer Deaths Per Capita by State in 2010",
       subtitle = "Color-coded map based on Cancer Deaths Per Capita (per 1,000 people)") +
  theme(legend.position = "bottom", legend.box = "horizontal")

top_cancer_data <- merged_data %>%
  filter(Cause.Name == "Cancer") %>%
  arrange(desc(Deaths_Per_Capita)) %>%
  top_n(10, Deaths_Per_Capita)

print(top_cancer_data[, c("State", "Deaths_Per_Capita")])
```

This map shows the deaths per 100,000 people for people who died because of suicide for each state. The lower table gives the numbers for the top 10 states.

```{r message=FALSE, echo=FALSE}
suicides_data <- total_data %>%
  filter(Year == 2010, Cause.Name == "Suicide") %>%
  select(State, Deaths, Age.adjusted.Death.Rate)

suicides_merged_data <- left_join(suicides_data, state_populations_2010, by = "State")

suicides_merged_data <- suicides_merged_data %>%
  mutate(Suicides_Per_Capita = Deaths / Population_2010 * 100000)  # Multiplying by 100,000 to get per 100,000 people

suicides_merged_data$state <- tolower(suicides_merged_data$State)

plot_usmap(data = suicides_merged_data, values = "Suicides_Per_Capita") +
  labs(title = "Suicides Per Capita by State",
       subtitle = "Color-coded map based on Suicides Per Capita (per 100,000 people)") +
  theme(legend.position = "bottom", legend.box = "horizontal")

top_suicide_data <- suicides_merged_data %>%
  top_n(10, Suicides_Per_Capita) %>%
  arrange(desc(Suicides_Per_Capita))

print(top_suicide_data[, c("State", "Suicides_Per_Capita")])

```

## Conclusion

I looked into the leading causes of death in the United States, seeking to determine patterns to help policymakers, healthcare professionals, and the public guide efforts to improve public health outcomes. By using a comprehensive data set from the National Center for Health Statistics and the U.S. Census, I used statistical techniques to analyze and interpret the leading causes of death. My approach includes investigating various death rate metrics and their implications for states and causes. My analysis has uncovered the top causes of death by number and state, shedding light on variations in mortality rates across the nation. Additionally, my exploration into age-adjusted death rates and per-capita death rates has provided nuanced perspectives on state-specific and cause-specific mortality patterns. The insights derived from my analysis can guide policymakers in prioritizing health interventions, assist healthcare professionals in targeted approaches, and empower the public with knowledge about health disparities and concerns in their respective states.
While my analysis provides valuable insights, it is not without limitations. The data's reliance on the 2010 census and the specific time frame may limit the representation of current health dynamics. Future analyses could benefit from more recent data and an exploration of additional factors influencing mortality. Also, including socio-economic variables and healthcare accessibility metrics could improve our understanding. Continuous refinement and expansion of data sets will contribute to more accurate and actionable insights.








