---
title: "How Are You Going to Die?"
author: "Andrew Zilnicki"
date: "2023-11-10"
output: 
  slidy_presentation:
    widescreen: true
    smaller: true
---

# Introduction

When thinking about death, do you ever wonder about what people actually
die from? <br> 
What are the causes that end lives? <br> 
Is there certain illnesses that affect more people?<br> 
Does which state you live in make a difference in how your life ends?

# Questions

**Domain Question**

What are the top causes of deaths by number and by state?

**Other Questions**

1.  Does using the Age-adjusted death rate show different information
    about deaths per state?
2.  Does using the Age-adjusted death rate show different information
    about deaths per cause?
3.  Does using the per-capita death rate show different information
    about deaths per state?
4.  In which states is cancer most deadly using the per-capita death rate?
5.  Which states have the highest rates of suicide using the per-capita death rate?

# Data Source

I got this from data.gov and the population data from the 2010 census.
<br> 
[Link to
Dataset](https://catalog.data.gov/dataset/nchs-leading-causes-of-death-united-states/resource/9096aa3c-0d4b-42f1-bb01-284816d92a15)

# Here are the packages I used

```{r message=FALSE}
library(DT)
library(usmap)
library(dplyr)
library(ggplot2)
```

# Unfiltered Data

This data set is from Data.gov. First, I had to download the data from
Data.gov. The data was downloaded via a CSV file. Once I created and R
Markdown file I uploaded the CSV file into a new data frame called
total_data.

```{r}
total_data <- read.csv("/Users/andrew/DATA 2401/Final project/NCHS_-_Leading_Causes_of_Death__United_States.csv")
glimpse(total_data)
```

## Data Key Terms:

**Year:** Year of death. <br> **X113 Cause Name:** Detailed name for
cause of death. <br> **Cause Name:** Short name for cause of death. <br>
**State:** State where death occurred. <br> **Deaths:** Total deaths for
that death and cause in a particular year. <br> **Age adjusted Death
Rate:** Formula determined by total expected number of deaths divided by
the sum of the standard population and the resulting quotient is
multiplied by 1,000,000. This is the death rate that would have existed
if the population under study had the same age distribution as the
"standard" population. This is useful for summary measures adjusted for
differences in age distributions.

# Data Analysis

To start analyzing data we group our data by state name and cause of
death. We can then plot the data of total deaths by each cause. To make
it easier to view we organize the causes from descending order.

```{r, echo=FALSE, message=FALSE}
result <- total_data %>%
  group_by(Cause.Name, State) %>%
  summarize(TotalDeaths = sum(Deaths)) 
```

## 

```{r, echo=FALSE}
result %>%
  filter(State == "United States" & Cause.Name != "All causes") %>%
  arrange(desc(TotalDeaths)) %>%
  ggplot(aes(x = reorder(Cause.Name, -TotalDeaths), y = TotalDeaths, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Deaths by Cause in the United States",
       x = "Cause Name",
       y = "Total Deaths",
       fill = "Cause Name") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 

Now filter our data by total deaths for each state ordering from the
least deaths to the most deathd for easy viewing.

```{r, echo=FALSE}
result %>%
  filter(Cause.Name == "All causes" & State != "United States") %>%
  arrange(desc(TotalDeaths)) %>%
  ggplot(aes(x = reorder(State, -TotalDeaths), y = TotalDeaths, fill = State)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flip the plot for a horizontal bar chart
  labs(title = "Total Deaths by State for All Causes",
       x = "Total Deaths",
       y = "State",
       fill = "State") +
  theme_minimal()
```

#

Now let's filter it down to the top 5 causes and the top 10 states.

```{r message=FALSE, echo=FALSE}
filtered_data <- result %>%
  filter(
    Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
    State %in% c("California", "Florida", "Texas", "New York", "Pennsylvania", "Ohio", "Illinois", "Michigan", "North Carolina", "New Jersey")
  )
```

```{r, echo=FALSE}
ggplot(filtered_data, aes(x = reorder(State, -TotalDeaths), y = TotalDeaths, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Deaths by Cause in Selected States",
       x = "Total Deaths",
       y = "State",
       fill = "Cause Name") +
  theme_minimal()
```

This shows that California has far more deaths than even the second
closest state of Florida. California has triple the number of deaths
from the top 5 causes once you look at North Carolina or New Jersey.

# A better way to look at causes?

It might look like some states have far more deaths from illness, but
these states also have the largest populations. <br>
What if we used the Age-adjusted data? <br>
Would we get different results for states and causes?

#

```{r, message=FALSE, echo=FALSE}
new_filtered_data <- total_data %>%
  filter(Cause.Name != "All causes" & State != "United States") %>%
  group_by(Cause.Name, State) %>%
  summarise(TotalAgeAdjustedDeathRate = sum(Age.adjusted.Death.Rate))
```

```{r, message=FALSE, echo=FALSE}
ggplot(new_filtered_data, aes(x = reorder(Cause.Name, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Age-Adjusted Death Rate by Cause Name",
       x = "Total Age-Adjusted Death Rate",
       y = "Cause Name",
       fill = "Cause Name") +
  theme_minimal()
```

#

```{r message=FALSE, echo=FALSE}
ggplot(new_filtered_data, aes(x = reorder(State, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = State)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Age-Adjusted Death Rate by State",
       x = "Total Age-Adjusted Death Rate",
       y = "State",
       fill = "State") +
  theme_minimal()
```

#

Now let's look at the top 5 Age-Adjusted Death Causes and the Top 10 states

```{r message=FALSE, echo=FALSE}
top_causes_states <- new_filtered_data %>%
  filter(Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
           State %in% c("Mississippi", "Kentucky", "Oklahoma", "West Virginia", "Alabama",
                         "Louisiana", "Arkansas", "Tennessee", "Missouri", "South Carolina"))

datatable(top_causes_states, options = list(pageLength = 10))
```

#

```{r message=FALSE, echo=FALSE}
ggplot(top_causes_states, aes(x = reorder(State, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 5 Age-Adjusted Death Causes by Top 10 States",
       x = "Total Age-Adjusted Death Rate",
       y = "State",
       fill = "Cause Name") +
  theme_minimal()
```

```{r message=FALSE, echo=FALSE}
result_question1 <- total_data %>%
  group_by(Cause.Name) %>%
  filter(!is.na(Deaths)) %>%
  slice(which.max(Deaths))

print(result_question1[, c("Year", "Cause.Name", "Deaths")])
```

# Using per capita

Let's look at the deaths using population per capita to see if that changes the results. <br>
I am going to pick 2010 and load population data from the census.

```{r message=FALSE}
# Creating a data frame with state populations from the 2010 Census
state_populations_2010 <- data.frame(
  State = c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado",
    "Connecticut", "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho",
    "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana",
    "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",
    "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey",
    "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma",
    "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
    "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia",
    "Wisconsin", "Wyoming"
  ),
  Population_2010 = c(
    4779736, 710231, 6392017, 2915918, 37253956, 5029196,
    3574097, 897934, 601723, 18801310, 9687653, 1360301, 1567582,
    12830632, 6483802, 3046355, 2853118, 4339367, 4533372,
    1328361, 5773552, 6547629, 9883640, 5303925, 2967297,
    5988927, 989415, 1826341, 2700551, 1316470, 8791894,
    2059179, 19378102, 9535483, 672591, 11536504, 3751351,
    3831074, 12702379, 1052567, 4625364, 814180, 6346105,
    25145561, 2763885, 625741, 8001024, 6724540, 1852994,
    5686986, 563626
  )
)
```

```{r message=FALSE, echo=FALSE}
total_data_2010 <- total_data %>% 
  filter(Year == 2010) %>% 
  select(Year, Cause.Name, State, Deaths, Age.adjusted.Death.Rate)

merged_data <- left_join(total_data_2010, state_populations_2010, by = "State")

merged_data <- merged_data %>%
  mutate(Deaths_Per_Capita = Deaths / Population_2010 * 1000)  # Multiplying by 1000 to get deaths per 1000 people
```

#
This map shows the deaths per 1,000 people for each state. The lower table gives the numbers for the top 10 states.

```{r message=FALSE, echo=FALSE}
merged_data$state <- tolower(merged_data$State)
plot_usmap(data = merged_data, values = "Deaths_Per_Capita", color = "lightblue") +
  labs(title = "Deaths Per Capita by State in 2010",
       subtitle = "Color-coded map based on Deaths Per Capita (per 1,000 people)") +
  theme(legend.position = "bottom", legend.box = "horizontal")

top_merged_data <- merged_data %>%
  filter(Cause.Name == "All causes") %>%
  arrange(desc(Deaths_Per_Capita)) %>%
  top_n(10, Deaths_Per_Capita)

print(top_merged_data[, c("State", "Deaths_Per_Capita")])
```

#
This map shows the deaths per 1,000 people for people who died because of cancer for each state. The lower table gives the numbers for the top 10 states.

```{r message=FALSE, echo=FALSE}
cancer_data <- merged_data %>%
  filter(Cause.Name == "Cancer")

cancer_data$state <- tolower(cancer_data$State)
  plot_usmap(data = cancer_data, values = "Deaths_Per_Capita", color = "lightblue") +
  labs(title = "Cancer Deaths Per Capita by State in 2010",
       subtitle = "Color-coded map based on Cancer Deaths Per Capita (per 1,000 people)") +
  theme(legend.position = "bottom", legend.box = "horizontal")

top_cancer_data <- merged_data %>%
  filter(Cause.Name == "Cancer") %>%
  arrange(desc(Deaths_Per_Capita)) %>%
  top_n(10, Deaths_Per_Capita)

print(top_cancer_data[, c("State", "Deaths_Per_Capita")])
```

#
This map shows the deaths per 100,000 people for people who died because of suicide for each state. The lower table gives the numbers for the top 10 states.

```{r message=FALSE, echo=FALSE}
suicides_data <- total_data %>%
  filter(Year == 2010, Cause.Name == "Suicide") %>%
  select(State, Deaths, Age.adjusted.Death.Rate)

suicides_merged_data <- left_join(suicides_data, state_populations_2010, by = "State")

suicides_merged_data <- suicides_merged_data %>%
  mutate(Suicides_Per_Capita = Deaths / Population_2010 * 100000)  # Multiplying by 100,000 to get per 100,000 people

suicides_merged_data$state <- tolower(suicides_merged_data$State)

plot_usmap(data = suicides_merged_data, values = "Suicides_Per_Capita") +
  labs(title = "Suicides Per Capita by State",
       subtitle = "Color-coded map based on Suicides Per Capita (per 100,000 people)") +
  theme(legend.position = "bottom", legend.box = "horizontal")

top_suicide_data <- suicides_merged_data %>%
  top_n(10, Suicides_Per_Capita) %>%
  arrange(desc(Suicides_Per_Capita))

print(top_suicide_data[, c("State", "Suicides_Per_Capita")])

```
