---
title: "Final Project"
author: "Andrew Zilnicki"
date: "2023-11-10"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Introduction

When thinking about death, do you ever wonder about what people actually die from? What are the causes that end lives? Is there certain illnesses that affect more people? Does which state you live in make a difference in how your life ends?

## Questions

**Domain Question**

What are the top causes of deaths by number and by state?

**Other Questions**

1. Which year had the most deaths by each cause of death?
2. Does using the Age-adjusted death rate show different information about deaths per state?
3. Does using the Age-adjusted death rate show different information about deaths per cause?
4. Does using the Age-adjusted death rate show different information about deaths per year?
5. In which states is cancer most deadly?
6. Which states have the highest amount of suicide both total and age-adjusted?

# Data Source

[Link to Dataset](https://catalog.data.gov/dataset/nchs-leading-causes-of-death-united-states/resource/9096aa3c-0d4b-42f1-bb01-284816d92a15)

# Get Ready

```{r}
library(dplyr)
library(ggplot2)
```

# Unfiltered Data
This data set is from Data.gov. First, I had to download the data from Data.gov. The data was downloaded via a CSV file. Once I created and R Markdown file I uploaded the CSV file into a new data frame called total_data.

```{r}
total_data <- read.csv("/Users/andrew/DATA 2401/Final project/NCHS_-_Leading_Causes_of_Death__United_States.csv")
glimpse(total_data)
```

## Data Key Terms:

**Year:** Year of death. <br>
**X113 Cause Name:** Detailed name for cause of death. <br>
**Cause Name:** Short name for cause of death. <br>
**State:** State where death occurred. <br>
**Deaths:** Total deaths for that death and cause in a particular year. <br>
**Age adjusted Death Rate:** Formula determined by total expected number of deaths divided by the sum of the standard population and the resulting quotient is multiplied by 1,000,000. This is the death rate that would have existed if the population under study had the same age distribution as the "standard" population. This is useful for summary measures adjusted for differences in age distributions.

## Data Analysis

To start analyzing data we group our data by state name and cause of death. We can then plot the data of total deaths by each cause. To make it easier to view we organize the causes from descending order.

```{r}
result <- total_data %>%
  group_by(Cause.Name, State) %>%
  summarize(TotalDeaths = sum(Deaths)) 
```

```{r}
result %>%
  filter(State == "United States" & Cause.Name != "All causes") %>%
  arrange(desc(TotalDeaths)) %>%
  ggplot(aes(x = reorder(Cause.Name, -TotalDeaths), y = TotalDeaths, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Deaths by Cause in the United States",
       x = "Cause Name",
       y = "Total Deaths",
       fill = "Cause Name") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We can next filter our data by total deaths for each state. Again, can order it from the most deaths to the least death for easier viewing.

```{r}
result %>%
  filter(Cause.Name == "All causes" & State != "United States") %>%
  arrange(desc(TotalDeaths)) %>%
  ggplot(aes(x = reorder(State, -TotalDeaths), y = TotalDeaths, fill = State)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flip the plot for a horizontal bar chart
  labs(title = "Total Deaths by State for All Causes",
       x = "Total Deaths",
       y = "State",
       fill = "State") +
  theme_minimal()
```

Now that we can see the total deaths per state and total per cause let's filter it down a little. Let's take the top 5 causes and the top 10 states.

```{r}
filtered_data <- result %>%
  filter(
    Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
    State %in% c("California", "Florida", "Texas", "New York", "Pennsylvania", "Ohio", "Illinois", "Michigan", "North Carolina", "New Jersey")
  )

print(filtered_data)
```

```{r}
ggplot(filtered_data, aes(x = reorder(State, -TotalDeaths), y = TotalDeaths, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Deaths by Cause in Selected States",
       x = "Total Deaths",
       y = "State",
       fill = "Cause Name") +
  theme_minimal()
```
<br> This shows that California has far more deaths than even the second closest state of Florida. California has triple the number of deaths from the top 5 causes once you look at North Carolina or New Jersey.

## A better way to look at causes?

It might look like some states have far more deaths from illness, but these states also have the largest populations. What if we used the Age-adjusted data? Would we get different results for states and causes?

```{r}
new_filtered_data <- total_data %>%
  filter(Cause.Name != "All causes" & State != "United States") %>%
  group_by(Cause.Name, State) %>%
  summarise(TotalAgeAdjustedDeathRate = sum(Age.adjusted.Death.Rate))

print(new_filtered_data)
```

```{r}
ggplot(new_filtered_data, aes(x = reorder(Cause.Name, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Age-Adjusted Death Rate by Cause Name",
       x = "Total Age-Adjusted Death Rate",
       y = "Cause Name",
       fill = "Cause Name") +
  theme_minimal()
```

```{r}
ggplot(new_filtered_data, aes(x = reorder(State, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = State)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Age-Adjusted Death Rate by State",
       x = "Total Age-Adjusted Death Rate",
       y = "State",
       fill = "State") +
  theme_minimal()
```
Now let's look at the top 5 Age-Adjusted Death Causes and the Top 10 states

```{r}
top_causes_states <- new_filtered_data %>%
  filter(Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
           State %in% c("Mississippi", "Kentucky", "Oklahoma", "West Virginia", "Alabama",
                         "Louisiana", "Arkansas", "Tennessee", "Missouri", "South Carolina"))

print(top_causes_states)
```

```{r}
ggplot(top_causes_states, aes(x = reorder(State, -TotalAgeAdjustedDeathRate), y = TotalAgeAdjustedDeathRate, fill = Cause.Name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 5 Age-Adjusted Death Causes by Top 10 States",
       x = "Total Age-Adjusted Death Rate",
       y = "State",
       fill = "Cause Name") +
  theme_minimal()
```

Let's compare the data from total deaths and age adjusted deaths.
```{r}
total_deaths_data <- total_data %>%
  filter(Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
           State %in% c("Mississippi", "Kentucky", "Oklahoma", "West Virginia", "Alabama",
                         "Louisiana", "Arkansas", "Tennessee", "Missouri", "South Carolina")) %>%
  group_by(Year, Cause.Name, State) %>%
  summarise(TotalDeaths = sum(Deaths))

age_adjusted_deaths_data <- total_data %>%
  filter(Cause.Name %in% c("Heart disease", "Cancer", "Stroke", "CLRD", "Unintentional injuries") &
           State %in% c("Mississippi", "Kentucky", "Oklahoma", "West Virginia", "Alabama",
                         "Louisiana", "Arkansas", "Tennessee", "Missouri", "South Carolina")) %>%
  group_by(Year, Cause.Name, State) %>%
  summarise(TotalAgeAdjustedDeathRate = sum(Age.adjusted.Death.Rate))

merged_data <- merge(total_deaths_data, age_adjusted_deaths_data, by = c("Year", "Cause.Name", "State"))

merged_data <- merged_data %>%
  group_by(State, Cause.Name) %>%
  summarise(TotalDeaths = sum(TotalDeaths),
            TotalAgeAdjustedDeathRate = sum(TotalAgeAdjustedDeathRate))
```

```{r}
ggplot(merged_data, aes(x = State, y = TotalDeaths, group = Cause.Name, color = "Total Deaths")) +
  geom_line(size = 1.5) +
  geom_line(aes(y = TotalAgeAdjustedDeathRate, color = "Age-Adjusted Deaths", linetype = "Age-Adjusted Deaths"), size = 1.5) +
  labs(title = "Total Deaths and Age-Adjusted Deaths by State",
       x = "State",
       y = "Count",
       color = "Legend") +
  theme_minimal() +
  theme(legend.position = "top", legend.title = element_blank())
```









